taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-12-11 paper/manuscript.Rmd
Beef up citations

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

tibble(provider = c("Integrated Taxonomic Information System, ITIS 2019",
                                 "National Center for Biological Information's Taxonomy database, NCBI",
                                 "Catalogue of Life, COL",
                                 "Global Biodiversity Information Facility Taxonomic Backbone, GBIF 2019",
                                 "FishBase, Froese and Pauly 2019",
                                 "Open Tree Taxonomy, OTT",
                                 "International Union for Conservation of Nature and Natural Resources,IUCN 2019"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

>>>>

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names. How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")
